---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  eval = FALSE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
library(knitr)
```

# tadaR
# R functions for processing and plotting output from TADA (TAsk Dynamics Application)

Sam Kirkham

### Loading data and accessing variables ###

Load tidyverse libraries (needed below)

```{r}
library(tidyverse)
```

Load a TADA file using 'readTADA' (here we use a synthesised production of the word 'pipe').

```{r}
d <- readTADA("~/Dropbox/projects/tardis/tada_modelling/tada_test/pipe_traj_mv.mat")
```

Now we can access columns as follows
```{r}
d$audio # all columns of 'audio' list
d$audio$SIGNAL # just the 'signal' column of 'audio'
plot(d$audio$SIGNAL, type = "l") # plot audio signal (need to add proper time variable)
```


### Plotting signals ###

We can plot tract variable signals (across the whole file)
```{r}
plot(d$TBCD$SIGNAL, type = "l") # TBCD
plot(d$TBCL$SIGNAL, type = "l") # TBCL
plot(d$TBCD$SIGNAL,d$TBCL$SIGNAL, type = "l") # TBCD~TBCL (what's this mean...? cool though)
```

Generally, we want one variable with just task variables, with one column per variable:
```{r}
d.wide <- unnestTADA(d)
```

The above plots are much easier to do using the now unnested data, as we can refer directly to variables:

```{r}
plot(d.wide$TBCL, type = "l") 
```

### Wide formant data ###

We can also convert the data to wide format. This allows us to easily show multiple variables on a single plot.

```{r}
d.long <- gather(d.wide, "Variable", "Value", -Sample, -Time)
d.long <- bind_rows(d.long, getAudioLong(d))
```

What articulators/variables are available to us?

```{r}
unique(d.long$Variable)
```

Plot selected variables over time using ggplot (i.e. Audio, Lip Aperture (LA), TBCL, TBCD, Glottis (GLO)).

```{r}
d.long %>% 
  filter(Variable %in% c("Audio", "LA", "TBCL", "TBCD", "GLO")) %>% 
  ggplot(aes(x = Time, y = Value)) +
  geom_path() +
  facet_wrap(~Variable, ncol = 1, scales = "free_y") + # let *only* y-scaling be free
  theme_minimal()
```


### Further notes ###

It would be nice to have segmental boundaries. However, TADA is not segmental, as it synthesises output based on overlapping gestures. Instead, we need to force-align the acoustic signal and obtain temporal landmarks from the forced-alignment.











